Constraints on magic state protocols from the statistical mechanics of Wigner negativity

### 2022-02-24 npj_revisions/manuscript files/magic_constraints_maintext.tex
final submission version

\hypersetup{
	colorlinks=true,  
	linkcolor=blue,   
	citecolor=blue,   
	urlcolor=blue     
}

\usepackage{lineno}
\linenumbers

@@

\subsection*{Overview of our results}
We begin by describing how stabilizer operations correspond to stochastic maps on a discrete phase space, and how majorization can be applied to constrain magic state transformations. We then consider families of magic protocols with increasing complexity. The simplest case we consider is for magic protocols that lead to unital channels. These are a subset of protocols that leave invariant some distinguished state or have some equilibrium fixed-point structure. We then consider protocols that are non-equilibrium processes, but only generate sub-linear correlations that enable a simple description in the thermodynamic limit before giving entropic analysis that can be applied generally.

The relations between the sets of protocols considered are as follows:
\begin{linenomath}\begin{align}
	\mbox{Unital} &\subset \mbox{ local fixed-points} \nonumber\\ 
&\subset \mbox{ sub-linear correlations} \subset \mbox{ general protocols.} \nonumber
\end{align}\end{linenomath}

@@

\begin{figure}[t]
    \centering
    \includegraphics[scale=0.38]{distill_bounds.pdf}
    \caption{\textbf{Magic distillation bounds from majorization of Wigner quasi-distributions.} The above plot shows distillation upper bounds for qutrits obtained from majorization on unital protocols. The constraints are plotted as a function of the depolarizing error $\epsilon$ for a noisy Strange state (Eq.~(\ref{eq:noisysn})). The bound $R_\infty$ follows from a Lorenz curve analysis presented before Eq.~(\ref{eq:unital-bound}), while the bound $R_{10}$ comes from single-shot entropies on the Wigner quasi-distributions, as discussed in the derivation of Eq.~(\ref{eq:Ra}). Numerical bounds $R_{\rm{num}}$ are obtained by using the full set of majorization constraints. The new bounds $R_\infty$, $R_{10}$, $R_{\rm{num}}$ all provide stricter constraints than mana~\cite{cit:veitch2} and max--thauma $\theta_{\rm{max}}$~\cite{Wang_2020} bounds for noisy Strange states.
    }
    \label{fig:distill_bounds}
\end{figure}

@@

\subsection*{Phase space representation of quantum mechanics}

@@

In the usual case where $\w$ and $\r$ are both probability distributions the Lorenz curve is defined on the interval $[0,1]$, and rises monotonically until it reaches the value $1$ at $x=1$. The value $L_{\w|\r}(x) = 1$ is a global maximum. Moreover, if $\w, \w', \r, \r'$ are all valid probability distributions with $\r,\r'$ having positive components, then
\begin{linenomath}\begin{equation*}
	(\w, \r) \succ (\w', \r') \mbox{ if and only if } L_{\w |\r}(x) \ge L_{\w' |\r'}(x),
\end{equation*}\end{linenomath}
for all $x \in [0,1]$. 
This result is shown in~\cite{ruch_mixing_1978} and we reproduce it in Supplementary Note~2.
It provides a simple way of computing whether relative majorization holds between pairs of probability distributions.

@@

We can now consider how majorization constrains magic distillation rates starting in this section with an abstract setting based on the fixed-point structure of the physical operations accessible to us. We show that any magic theory can always be decomposed in terms of `sub-theories' with a particular fixed-point structure, which allows the use of $\d$--majorization. 
Later, we drop the requirement of fixed-points and instead use the full relative majorization, before giving general entropic constraints.

@@

We consider some $\sigma \in \F$, and its corresponding $\R_\sigma$. We are interested in the ability to transform many copies of some noisy magic state $\rho$ towards a more pure form of magic. The state $\rho$ is assumed to have negativity in the Wigner representation, so $W_\rho(\z) < 0$ for some regions of $\z \in \P_d$. The state $\sigma$ is assumed to have a Wigner distribution with $W_\sigma(\z) > 0$ for all $\z$ in the phase space.
This is justified because a non-full rank state $\sigma$ can be handled as a limiting case in which we first add an infinitesimal fraction of depolarising noise $\epsilon (\id/d)$ and then take $\epsilon \rightarrow 0$.

@@

Before computing explicit examples, we state some generic aspects of Lorenz curves for magic states, which allow us to interpret previous magic monotones in a new light. The first result gives a simple way to see that the sum-negativity/mana of a magic state is a monotone~\cite{cit:veitch2}.

@@

This figure includes numerical estimates of the optimal majorization bounds $R_{\rm{num}}$ obtained by considering the full Lorenz curve, which show that the bound $R_\infty$ can be further improved. While this is an improvement on prior results,  all known distillation protocols have rates much lower than these upper bounds.  It remains a major open question to determine what are the best possible rates that can be achieved.

@@

Then the asymptotic magic distillation rate $R = m/n$ is bounded as
\begin{equation}\label{eq:rate_bounds_proof}
	R \leq \dfrac{\log \big( 1-\frac{4}{3}\epsilon \big) + \beta (\phi - F)}{\log \big( 1-\frac{4}{3}\epsilon' \big) + \beta (\phi' - F')},
\end{equation}
where $F$ is the free energy of $\tau$,  and 
\begin{equation}\label{eq:phi}
	\phi = -\beta^{-1} \log \zeta
\end{equation}
with $\zeta$ given by the expressions
\begin{linenomath}\begin{align}
	\zeta &= \sum_{k\in \mathbb{Z}_3} \alpha_k e^{-\beta E_k}, \\
	\alpha_k &= \<E_k| A_{\z_\star} |E_k\>,
\end{align}\end{linenomath}
and $W_\tau(\z)$ attaining a minimum at $\z=\z_\star$. The primed variables are defined similarly for the output system.
\end{theorem}
\noindent The proof of this is provided in Supplementary Note~4, and follows a similar line to the unital protocol bound.

@@

\subsection*{Technical aspects and special cases of free energy bounds}

@@

The $\alpha \rightarrow 1$ limit of the R\'{e}nyi entropy diverges if negativity is present, however it diverges in a well-defined way. We can write the entropy as 
\begin{equation}
	H_{1+\epsilon}(W_\rho) = -\epsilon^{-1} \log \sum_{\z} |W_\rho (\z)|^{1+\epsilon}
\end{equation} 
with $\epsilon \rightarrow 0^+$ through a sequence of rational values. We recall that mana can be written as $\M(\rho) = \log \sum_{\z} |W_\rho(\z)|$, so we obtain a second new perspective on mana:
\begin{equation}\label{eq:mana-divergence}
\M(\rho) = -\lim_{\epsilon \rightarrow 0^+} \epsilon H_{1+\epsilon}(W_\rho).
\end{equation}
Therefore, the mana of a state $\rho$ is minus the residue of the pole at $\alpha=1$ for $H_\alpha(W_\rho)$, and quantifies the divergence of the R\'{e}nyi-entropy as we approach the limiting Shannon entropy $H(W_\rho)$.

@@

Because Theorem \ref{thm:frag}  provides a completeness statement for majorization, we can ask if this approach could additionally provide \emph{lower bounds} for distillation protocols. In contrast to the upper bounds, it is now essential to include the symplectic constraints on the phase space in the majorization relations. In principle the resultant majorization (and relative majorization) constraints should provide exact specifications on what can be obtained via free operations. 
Interestingly, this route does not require the use of error correcting codes, but instead is built on group-theoretic features of the phase space. 
The Clifford group action on a quantum system corresponds to the action of the affine symplectic group $G \coloneqq SL(2,\mathbb{Z}_d) \ltimes \mathbb{Z}_d^2$ on the discrete phase space~\cite{Appleby_2005, Bengtsson_2006}. Therefore, any convex mixture of Clifford unitaries will correspond to a convex mixture of these group actions. The Hardy-Littlewood theorem~\cite{hardy_1952} tells us that majorization is obtained from convex mixtures of arbitrary permutations, a result that has also been generalized to convex mixtures of elements of a general group $G$. $G$--majorization has been studied in the classical literature and a range of results are known about it~\cite{giovagnoli_1985, steerneman_1990, giovagnoli_cyclic_1996, eaton_1977}. 
One can therefore consider sub-groups of the affine symplectic group and analyse the resulting majorization/relative majorization that follows from its action on quasi-distributions. However if the sub-group considered is too simple then it may lead to trivial distillation rates. An example of this is when we restrict to the set of Weyl covariant channels~\cite{fukuda2006}, which are represented by a convex mixture of displacements on the phase space. Initial work by the authors have shown that the majorization conditions for Weyl covariant channels can be solved exactly in terms of discrete Fourier transforms, however it is found that the resultant distillation rates are trivial.

One notable structure that could be exploited for $G$--majorization is when $G$ is a finite reflection group. For this,  the $G$--majorization pre-order is guaranteed to be described by a finite list of conditions~\cite{giovagnoli_1985}, just as is the case for the relative majorization ordering. Therefore one route to concrete lower bounds on distillation rates is to consider stochastic maps obtained by reflection sub-groups of $SL(2,\mathbb{Z}_d) \ltimes \mathbb{Z}_2$, or other simple sub-groups, for which the majorization relation reduces to a finite set of conditions. Beyond this, another promising route is to formulate the majorization conditions for a sub-set of operations as a semi-definite program and then attempt to estimate realisable distillation rates by exploiting dualities and appropriate coarse-grainings. This approach has recently been applied by one of the authors to estimate the set of quantum states accessible by covariant channels in the resource theory of asymmetry~\cite{alexander2021} where one allows depolarizing noise that is then minimized. 
It would be of interest to see if a similar approach could be followed for magic state distillation.

@@

\section*{Methods}

@@

Precise definitions of all the above concepts and explanations of their incorporation to the study of magic distillation are provided in the Results section as the concepts become relevant.

@@

\bibitem{cit:veitch2}
\bibinfo{author}{Veitch, V.}, \bibinfo{author}{Mousavian, S. A.~H.},
  \bibinfo{author}{Gottesman, D.} \& \bibinfo{author}{Emerson, J.}
\newblock \bibinfo{title}{The resource theory of stabilizer quantum
  computation}.
\newblock \emph{\bibinfo{journal}{New Journal of Physics}}
  \textbf{\bibinfo{volume}{16}}, \bibinfo{pages}{013009}
  (\bibinfo{year}{2014}).
\newblock \urlprefix\url{https://doi.org/10.1088/F1367-2630/16/1/013009}.

>>>>

\newcommand{\revhigh}[1]{{\color{red}#1}}
\hypersetup{
	colorlinks=true,  
	linkcolor=blue,   
	citecolor=blue,   
	urlcolor=blue     
}

@@

\subsection*{Overview of our results}
\revhigh{
We begin by describing how stabilizer operations correspond to stochastic maps on a discrete phase space, and how majorization can be applied to constrain magic state transformations. We then consider families of magic protocols with increasing complexity. The simplest case we consider is for magic protocols that lead to unital channels. These are a subset of protocols that leave invariant some distinguished state or have some equilibrium fixed-point structure. We then consider protocols that are non-equilibrium processes, but only generate sub-linear correlations that enable a simple description in the thermodynamic limit before giving entropic analysis that can be applied generally.}

The relations between the sets of protocols considered are as follows:
\begin{align}
\mbox{Unital} &\subset \mbox{ local fixed-points} \nonumber\\ 
&\subset \mbox{ sub-linear correlations} \subset \mbox{ general protocols.} \nonumber
\end{align}

@@

\begin{figure}[t]
    \centering
    \includegraphics[scale=0.38]{distill_bounds.pdf}
    \caption{\textbf{Magic distillation bounds from majorization of Wigner quasi-distributions.} The above plot shows distillation upper bounds for qutrits obtained from majorization on unital protocols. The constraints are plotted as a function of the depolarizing error $\epsilon$ for a noisy Strange state (Eq.~(\ref{eq:noisysn})). \revhigh{The bound $R_\infty$ follows from a Lorenz curve analysis presented before Eq.~(\ref{eq:unital-bound}), while the bound $R_{10}$ comes from single-shot entropies on the Wigner quasi-distributions, as discussed in the derivation of Eq.~(\ref{eq:Ra})}. Numerical bounds $R_{\rm{num}}$ are obtained by using the full set of majorization constraints. The new bounds $R_\infty$, $R_{10}$, $R_{\rm{num}}$ all provide stricter constraints than mana~\cite{cit:veitch2} and max--thauma $\theta_{\rm{max}}$~\cite{Wang_2020} bounds for noisy Strange states.
    }
    \label{fig:distill_bounds}
\end{figure}

@@

\revhigh{
\subsection*{Phase space representation of quantum mechanics}
}

@@

In the usual case where $\w$ and $\r$ are both probability distributions the Lorenz curve is defined on the interval $[0,1]$, and rises monotonically until it reaches the value $1$ at $x=1$. The value $L_{\w|\r}(x) = 1$ is a global maximum. Moreover, if $\w, \w', \r, \r'$ are all valid probability distributions with $\r,\r'$ having positive components, then
\begin{equation*}
(\w, \r) \succ (\w', \r') \mbox{ if and only if } L_{\w |\r}(x) \ge L_{\w' |\r'}(x),
\end{equation*}
for all $x \in [0,1]$. 
\revhigh{This result is shown in~\cite{ruch_mixing_1978} and we reproduce it in Supplementary Note~2.}
It provides a simple way of computing whether relative majorization holds between pairs of probability distributions.

@@

We can now consider how majorization constrains magic distillation rates starting in this section with an abstract setting based on the fixed-point structure of the physical operations accessible to us. We show that any magic theory can always be decomposed in terms of `sub-theories' with a particular fixed-point structure, which allows the use of $\d$--majorization. 
\revhigh{Later, we drop the requirement of fixed-points and instead use the full relative majorization, before giving general entropic constraints.}

@@

We consider some $\sigma \in \F$, and its corresponding $\R_\sigma$. We are interested in the ability to transform many copies of some noisy magic state $\rho$ towards a more pure form of magic. The state $\rho$ is assumed to have negativity in the Wigner representation, so $W_\rho(\z) < 0$ for some regions of $\z \in \P_d$. The state $\sigma$ is assumed to have a Wigner distribution with $W_\sigma(\z) > 0$ for all $\z$ in the phase space.
\revhigh{This is justified because a non-full rank state $\sigma$ can be handled as a limiting case in which we first add an infinitesimal fraction of depolarising noise $\epsilon (\id/d)$ and then take $\epsilon \rightarrow 0$.}

@@

\revhigh{Before computing explicit examples,} we state some generic aspects of Lorenz curves for magic states, which allow us to interpret previous magic monotones in a new light. The first result gives a simple way to see that the sum-negativity/mana of a magic state is a monotone~\cite{cit:veitch2}.

@@

This figure includes numerical estimates of the optimal majorization bounds $R_{\rm{num}}$ obtained by considering the full Lorenz curve, which show that the bound $R_\infty$ can be further improved. \revhigh{While this is an improvement on prior results,  all known distillation protocols have rates much lower than these upper bounds.  It remains a major open question to determine what are the best possible rates that can be achieved.}

@@

Then the asymptotic magic distillation rate $R = m/n$ is bounded as
\begin{equation}\label{eq:rate_bounds_proof}
	R \leq \dfrac{\log \big( 1-\frac{4}{3}\epsilon \big) + \beta (\phi - F)}{\log \big( 1-\frac{4}{3}\epsilon' \big) + \beta (\phi' - F')},
\end{equation}
where $F$ is the free energy of $\tau$,  and 
\begin{equation}\label{eq:phi}
	\phi = -\beta^{-1} \log \zeta
\end{equation}
with $\zeta$ given by the expressions
\begin{align}
	\zeta &= \sum_{k\in \mathbb{Z}_3} \alpha_k e^{-\beta E_k}, \\
	\alpha_k &= \<E_k| A_{\z_\star} |E_k\>,
\end{align}
and $W_\tau(\z)$ attaining a minimum at $\z=\z_\star$. The primed variables are defined similarly for the output system.
\end{theorem}
\noindent The proof of this is provided in Supplementary Note~4, and follows a similar line to the unital protocol bound.

@@

\revhigh{
\subsection*{Technical aspects and special cases of free energy bounds}
}

@@

The $\alpha \rightarrow 1$ limit of the R\'{e}nyi entropy diverges if negativity is present, however it diverges in a well-defined way. We can write the entropy as 
\begin{equation}
	H_{1+\epsilon}(W_\rho) = -\epsilon^{-1} \log \sum_{\z} |W_\rho (\z)|^{1+\epsilon}
\end{equation} 
with $\epsilon \rightarrow 0^+$ through a sequence of rational values. We recall that mana can be written as $\M(\rho) = \log \sum_{\z} |W_\rho(\z)|$, so we obtain a second new perspective on mana:
\begin{equation}\label{eq:mana-divergence}
	\M(\rho) = -\lim_{\epsilon \rightarrow 0^+} \epsilon H_{1+\epsilon}(W_\rho).
\end{equation}
Therefore, the mana of a state $\rho$ is minus the residue of the pole at $\alpha=1$ for $H_\alpha(W_\rho)$, and quantifies the divergence of the R\'{e}nyi entropy as we approach the limiting Shannon entropy $H(W_\rho)$.

@@

Because Theorem \ref{thm:frag}  provides a completeness statement for majorization, we can ask if this approach could additionally provide \emph{lower bounds} for distillation protocols. In contrast to the upper bounds, it is now essential to include the symplectic constraints on the phase space in the majorization relations.  \revhigh{In principle the resultant majorization (and relative majorization) constraints should provide exact specifications on what can be obtained via free operations. 
Interestingly, this route does not require the use of error correcting codes, but instead is built on group-theoretic features of the phase space. } The Clifford group action on a quantum system corresponds to the action of the affine symplectic group $G \coloneqq SL(2,\mathbb{Z}_d) \ltimes \mathbb{Z}_d^2$ on the discrete phase space~\cite{Appleby_2005, Bengtsson_2006}. Therefore, any convex mixture of Clifford unitaries will correspond to a convex mixture of these group actions. The Hardy-Littlewood theorem~\cite{hardy_1952} tells us that majorization is obtained from convex mixtures of arbitrary permutations, a result that has also been generalized to convex mixtures of elements of a general group $G$. $G$--majorization has been studied in the classical literature and a range of results are known about it~\cite{giovagnoli_1985, steerneman_1990, giovagnoli_cyclic_1996, eaton_1977}. 
\revhigh{One can therefore consider sub-groups of the affine symplectic group and analyse the resulting majorization/relative majorization that follows from its action on quasi-distributions. However if the sub-group considered is too simple then it may lead to trivial distillation rates. An example of this is when we restrict to the set of Weyl covariant channels~\cite{fukuda2006}, which are represented by a convex mixture of displacements on the phase space. Initial work by the authors have shown that the majorization conditions for Weyl covariant channels can be solved exactly in terms of discrete Fourier transforms, however it is found that the resultant distillation rates are trivial.

One notable structure that could be exploited for $G$--majorization is when $G$ is a finite reflection group. For this,  the $G$--majorization pre-order is guaranteed to be described by a finite list of conditions~\cite{giovagnoli_1985}, just as is the case for the relative majorization ordering.} Therefore one route to concrete lower bounds on distillation rates is to consider stochastic maps obtained by reflection sub-groups of $SL(2,\mathbb{Z}_d) \ltimes \mathbb{Z}_2$, or other simple sub-groups, for which the majorization relation reduces to a finite set of conditions. \revhigh{Beyond this, another promising route is to formulate the majorization conditions for a sub-set of operations as a semi-definite program and then attempt to estimate realisable distillation rates by exploiting dualities and appropriate coarse-grainings. This approach has recently been applied by one of the authors to estimate the set of quantum states accessible by covariant channels in the resource theory of asymmetry~\cite{alexander2021} where one allows depolarizing noise that is then minimized. 
It would be of interest to see if a similar approach could be followed for magic state distillation.}

@@

\revhigh{
\section*{Methods}

@@

Precise definitions of all the above concepts and explanations of their incorporation to the study of magic distillation are provided in the Results section as the concepts become relevant.
}

@@

\bibitem{cit:veitch2}
\bibinfo{author}{Veitch, V.}, \bibinfo{author}{Mousavian, S. A.~H.},
  \bibinfo{author}{Gottesman, D.} \& \bibinfo{author}{Emerson, J.}
\newblock \bibinfo{title}{The resource theory of stabilizer quantum
  computation}.
\newblock \emph{\bibinfo{journal}{New Journal of Physics}}
  \textbf{\bibinfo{volume}{16}}, \bibinfo{pages}{013009}
  (\bibinfo{year}{2014}).
\newblock \urlprefix\url{https://doi.org/10.1088/1367-2630/16/1/013009}.
